Natural Selection in the 20th Century

### 2020-11-13 why-negative-selection.Rmd
Add discussion of changes between generations.

The parents' generation has a further source of ascertainment bias: sampling
parents of respondents overweights parents who have many children. For instance,
parents of three children will have, on average, three times more children
represented in UK Biobank than parents of one child. Parents of no children will
by definition not be represented. To compensate, we reweight our preferred
weightings (Age/Qualification) by the inverse of *number of siblings*. We then
regress this on our polygenic scores. Selection effects are highly correlated across
the two generations and most share the same sign (appendix Figure \@ref(fig:plot-res-sibs)). Effect sizes appear larger in the parents' generation. 
We treat this result cautiously, because when we
split respondents up by year of birth, we find few differences in effect
sizes between early- and late-born respondents, for either generation. In other
words, since estimated effect sizes change when we change the dependent
variable, but do not change over time within either dependent variable,
the explanation may be due to remaining ascertainment bias within our
sample.

Why do selection effects have *opposite* signs across older and younger mothers
at first live birth, and opposite signs when AFLB is controlled for? A possible
answer is given by the economic theory of fertility 
[@becker1960economic; @willis1973new; @becker1976child].
According to this, increases in potential earned income affect fertility via two
opposing channels. There is an "income effect" by which children become more
affordable, like any other good. There is also a "substitution effect": since
childrearing has a cost in time, the opportunity cost of childrearing increases
if one's market wage is higher. The income effect would lead higher earners to
have more children. The substitution effect would lead higher earners to have
fewer children. If so, then genetic variants which affect earnings
potential in the labour market may cause opposing effects on
fertility. The income effect will cause natural selection in favour of 
earnings-increasing variants. The substitution effect will do the reverse.

This theory predicts that genetic variants will be selected for (or against) in
proportion to their effect on earnings. Figure
\@ref(fig:plot-cors-earnings-educ) plots selection effects on each PGS against
that PGS's correlation with earnings in a respondent's first job[^earnings-est],
and against its correlation with educational attainment, a predictor of lifetime
earnings. The raw relationships (left column) are strongly negative. If we plot
selection effects controlling for age at first live birth (right column), the
effect reverses and becomes positive. This is evidence that the labour market
plays an important role in natural selection, with substitution effects
dominating overall.

@@

Suppose that people are aware of their endowments of human capital, and 
that education and innate human capital are complements. They choose how long
to stay in education. After leaving education they enter the labour market and
form families. If so, then people with less human capital will leave education 
earlier and have their first child earlier. Whilst in the labour market, higher
earners will have more children (i.e. income effects dominate). These 
assumptions will lead to the pattern we observe where earnings-increasing
variants lead to later first births (and fewer children overall), but more
children for any given AFLB.

@@

@@

Empirical work shows similar relationships between *earnings* or *education* and
fertility, to the relationships shown here between *genetics* and fertility.
Higher income is associated with lower fertility [@becker1960economic;
@jones2006economic].
@cohen2013financial find that income decreases fertility at low income levels
but increases it at higher income levels. In a US sample, fertility decreases
with education faster among single mothers than married mothers 
[@baudin2015fertility].

The economic theory predicts that natural selection on polygenic scores is mediatet
by labour market earnings. As a final test of this,
we re-estimate equation (\@ref(eq:regression)) for each PGS, 
controlling for earnings in first job and education levels. Effect sizes are generally reduced (appendix Figure \@ref(fig:plot-res-earn-educ)), and only `r n_05_33` 
out of 33 PGS are significant at $p = 0.05/33$.

@@

Polygenic scores could capture effects that are really due to population
stratification, although would not change our results for natural
selection of the scores. In \@ref(fig:plot-regs-controlled) we show
results for selection on polygenic scores residualized for the top 100
principal components of the genetic data, calculated within the UK
Biobank population. (TODO Abdel details.)

@@

```{r plot-regs-controlled, fig.cap = "Effects of residualized polygenic scores on number of siblings/children."}

@@

```{r plot-age-birth-parents, fig.cap = "Effect sizes on number of siblings controlling for parents' age at birth, eldest siblings"}

>>>>

The parents' generation has a further source of ascertainment bias: sampling
parents of respondents overweights parents who have many children. For instance,
parents of three children will have, on average, three times more children
represented in UK Biobank than parents of one child. Parents of no children will
by definition not be represented. To compensate, we reweight our preferred
weightings (Age/Qualification) by the inverse of *number of siblings*. We then
regress this on our polygenic scores. Selection effects are highly correlated
across the two generations, and most share the same sign (appendix Figure
\@ref(fig:plot-res-sibs)). Effect sizes appear larger in the parents'
generation. We treat this result cautiously, because when we split respondents
up by year of birth, we find few differences in effect sizes between early- and
late-born respondents, for either generation. In other words, since estimated
effect sizes change when we change the dependent variable, but do not change
over time within either dependent variable, the explanation may be due to
remaining ascertainment bias within our sample.

Although the direction of selection effects does not change between the generations,
there are some differences. In particular, after we residualize polygenic scores
on the first 100 principal components of the array data, selection effects on 
*number of siblings* are reduced by about ten percent on average, whereas effects on
*number of children* barely change (appendix Figure 
\@ref(fig:plot-regs-controlled-pcs)). One possible explanation for this is that
earlier fertility is driven more by poverty, which is geographically clustered
and therefore correlates with the broad-scale genetic variation captured by
principal components. (TODO: insurance motive.) Another difference is that
in the parents' generation, raw selection effects and effects controlling for
AFLB are positively correlated, whereas they are negatively correlated in
the respondents' generation.[^educ-exception] These differences suggest that
the mechanisms of natural selection may have changed over time, even though
the outcomes are broadly the same.

[^educ-exception] Educational attainment is an exception. It is selected 
against in the parents' generation, but selected for if we control for AFLB,
which is the same pattern as for respondents.

Our empirical results can be explained by the economic
theory of fertility [@becker1960economic; @willis1973new; @becker1976child].
According to this, increases in a person's wage affect their fertility via two
opposing channels. There is an *income effect* by which children become more
affordable, like any other good. There is also a *substitution effect*: since
childrearing has a cost in time, the opportunity cost of childrearing increases
if one's market wage is higher. The income effect would lead higher earners to
have more children. The substitution effect would lead higher earners to have
fewer children. If so, then genetic variants which affect earnings potential in
the labour market may cause opposing effects on fertility. The income effect
will cause natural selection in favour of earnings-increasing variants. The
substitution effect will do the reverse.

This theory predicts that genetic variants will be selected for (or against) in
proportion to their effect on earnings. Figure
\@ref(fig:plot-cors-earnings-educ) plots selection effects on each PGS against
that PGS's correlation with earnings in a respondent's first job[^earnings-est],
and against its correlation with educational attainment, a predictor of lifetime
earnings. The raw relationships (left column) are strongly negative. If we plot
selection effects controlling for age at first live birth (right column), the
effect reverses and becomes positive. This is evidence that the labour market
plays an important role in natural selection, with substitution effects
dominating overall. We also re-estimate equation \@ref(eq:regression) for each
PGS, controlling for earnings in first job and education levels. Effect sizes
are generally reduced (appendix Figure \@ref(fig:plot-res-earn-educ)), and only
`r n_05_33` out of 33 PGS are significant at $p = 0.05/33$, suggesting that
earnings and education indeed mediate the effect of PGS on fertility.

@@

This theory can explain why selection effects have opposite signs when AFLB
is controlled for. Suppose that people are aware of their endowments of human
capital, and that education and innate human capital are complements. They
choose how long to stay in education. After leaving education they enter the
labour market and form families. If so, then people with less human capital will
leave education earlier and have their first child earlier. This is a pure
substitution effect: the income effect plays no role in the decision to leave
education, since people are not yet earning wages. Whilst in the labour
market, higher earners will have more children (i.e. income effects dominate).
These assumptions will lead to the pattern we observe: earnings-increasing
variants lead to later first births (and fewer children overall), but more
children for any given AFLB.

@@

* TODO: more.

@@

Previous work shows similar relationships between *earnings* or *education* and
fertility, to the relationships shown here between *genetics* and fertility.
Higher income is associated with lower fertility [@becker1960economic;
@jones2006economic].
@cohen2013financial find that income decreases fertility at low income levels
but increases it at higher income levels. In a US sample, fertility decreases
with education faster among single mothers than married mothers 
[@baudin2015fertility].

@@

Polygenic scores could capture effects that are really due to population
stratification, although would not change our results for natural
selection of the scores. Figure \@ref(fig:plot-regs-controlled-pcs) shows
selection effects on polygenic scores residualized for the top 100
principal components of the genetic data, calculated within the UK
Biobank population.

@@

```{r plot-regs-controlled-pcs, fig.cap = "Effects of residualized polygenic scores on number of siblings/children."}

@@

```{r plot-age-birth-parents, fig.cap = "Selection effects controlling for parents' age at birth, eldest siblings"}
